---
title: "ENVS 193DS - Final"
author: Scott Wagner
date: "June 4, 2025"
format: html
toc: true
---

# Github Repository
https://github.com/ScottWag/ENVS-193DS_spring-2025_final

# Setup
```{r}
#| message: false
#| warning: false

#Load in packages
library(tidyverse)
library(here)
library(janitor)
library(dplyr)
library(ggplot2)
library(DHARMa)
library(MuMIn)
library(ggeffects)

#Read in sea surface temperature data
sst <- read_csv(here("data", "SST_update2023.csv"))
#read in data for problem 3 
nest_boxes <- read_csv(here("data", "occdist.csv"))

```
# Problem 1. Research writing

## a. Transparent statistical methods

In part 1, they used a Pearson correlation to test the linear relationship between the two variables, distance from headwater and nitrogen load. In part 2, they used a one-way ANOVA to test for differences in average nitrogen load among the six different nitrogen sources.

## b. More information needed

Two additional tests/pieces of information that should be included to provide more context for the results would be a test for effect size (eta-squared) and a data visualization (boxplot or bar chart). An effect size test tells the viewer whether the observed differences are statistically significant, represented by a p-value. A data visualization helps communicate the patterns, variability, and group differences in a more accessible way by making the analysis visible.

## c. Suggestions for rewriting

# Problem 2. Data visualization

## a. Cleaning and summarizing

```{r}

#Clean the data
sst_clean <- sst %>%
  mutate(
    date = ymd(date), # converts the 'date' column to date format 
    year_num = year(date), # extracts the numeric year from the date
    month_num = month(date), # extracts the numeric month from the date
    month = factor(substr(month.name[month_num], 1, 3), levels = substr(month.name, 1, 3), ordered = TRUE) # converts months to three letter ordered factor 
  ) %>%
  filter(year_num >= 2018 & year_num <= 2023) %>% # only include years 2018-2023
  mutate(year = as.factor(year_num)) %>% #converts numeric year to factor 
  group_by(year, month) %>% # groups by year and month for monthly averages
  summarise(mean_monthly_sst = round(mean(temp, na.rm = TRUE), 1), .groups = "drop") # computes mean sst and round to 1 decimal, ungroup after summarizing 

#Display 5 rows
slice_sample(sst_clean, n =5)
#Display the structure of cleaned data
str(sst_clean)

```

## b. Visualize the data

```{r}
#| message: false
#| warning: false

#Create visual in ggplot
ggplot(sst_clean, aes(x = month, y = mean_monthly_sst, group = year, color = year)) +
  geom_line(linewidth = 1) + # adds lines for each year
  geom_point(size = 2) + # adds dots for each monthly SST value
  scale_y_continuous(limits = c(12.4,20)) + # custom y-axis values
  labs(
    x = "Month", # x-axis label
    y = "Mean monthly sea surface temperature (°C)", # y-axis label
    color = "Year" # legend title
  ) +
  theme_minimal(base_size = 14) + # clean, minimal theme with text size
  theme(
    legend.position = c(0.1, 0.7), # places legend on plot with location
    axis.text.x = element_text(angle = 0), # ensures month labels are horizontal
    panel.grid.minor = element_blank(), # removes major gridlines
    panel.grid.major = element_blank(), # removes minor gridlines
    panel.border = element_rect(color = "black", fill = NA, size = 1) # adds black border
  ) +
  scale_color_brewer(palette = "Blues") # applies sequential blue color by year
```

# Problem 3. Data analysis

```{r}
#| include: false

# creating data frame
nest_boxes_clean <- nest_boxes %>%
  #clean names
  clean_names() %>%
  # add factors for modeling
  mutate(
    season = factor(season),
    repeated_use = factor(repeated_use),
    box_occupant = factor(box_occupant)
  )

glimpse(nest_boxes_clean)  

```


## a. Response variable

In the data set, 1s represent nest boxes occupied by Swift Parrots, the target species of the study. 0s represent boxes that were not occupied by Swift Parrots, meaning they either did not use it, or were out competed by another species. 

## b. Purpose of study

The study compares nest box usage between Swift Parrots, Common Starlings, and Tree Martins. The main difference is that Swift Parrots, the conservation target species, are critically endangered and the study goal is to understand whether the nest boxes help the target species or benefit its competitors. 

## c. Difference in “seasons”

The difference in "seasons" refers to the years 2016 and 2019, when the box occupancy data was collected. The years differ in terms of the population dynamics, environmental conditions, and competitive pressures of the two other birds.

## d. Table of models

| Model number | Season | Distance to forest edge | Model description | 
|:------------:|:------:|:-----------------------:|:-----------------:|  
| 0            |        |                         | Null model        | 
| 1            |    X   |            X            | Saturated model   |  
| 2            |    X   |                         | Season only       | 
| 3            |        |            X            | Distance from forest edge only|     


## e. Run the models

```{r}
# Run models
 
# model 0: Null model
model0 <- glm(sp ~ 1, family = binomial, data = nest_boxes_clean)
# model 1: Saturated model
model1 <- glm(sp ~ season + edge_distance, family = binomial, data = nest_boxes_clean)
# model 2: Season only 
model2 <- glm(sp ~ season, family = binomial, data = nest_boxes_clean)
# model 3: Distance from forest edge only 
model3 <- glm(sp ~ edge_distance, family = binomial, data = nest_boxes_clean)

```


## f. Check the diagnostics

```{r}
# check for diagnostics using DHARMa package
 
# For model 0
residual0 <- simulateResiduals(model0) 
plot(residual0)

# For model 1
residual1 <- simulateResiduals(model1) 
plot(residual1)

# For model 2
residual2 <- simulateResiduals(model2) 
plot(residual2)

# For model 3
residual3 <- simulateResiduals(model3) 
plot(residual3)

```


## g. Select the best model

```{r}
# use MuMIn pakage to select the best model through AIC
model.sel(model0, model1, model2, model3)
```
The best model as determined by Akaike's Information Criterion (AIC) was the saturated model, including both season and distance to forest edge.

## h. Visualize model predictions

```{r}

# get predictions from model1 using ggpredict
prediction <- ggpredict(model1, terms = c("edge_distance [0:1100 by=10]", "season"))

# plot predictions
ggplot(prediction, aes(x = x, y = predicted, color = group)) +
  geom_line(linewidth = 1.2) +  # predicted line
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.25, color = NA) +  # 95% confidence interval
  geom_jitter(data = nest_boxes_clean, aes(x = edge_distance, y = sp, color = season),
              height = 0.05, width = 15, alpha = 0.4, inherit.aes = FALSE) +  # data points
  labs(
    x = "Distance from Forest Edge (m)", # x-axis label
    y = "Probability of Swift Parrot Occupancy", #y-axis label
    color = "Season", # color by season
    fill = "Season" # fill by season
  ) +
  theme_minimal(base_size = 14) +
  theme(panel.grid = element_blank()) +  # remove gridlines
  scale_color_manual(values = c("2016" = "lightblue", "2019" = "seagreen")) +  # custom colors (lines/points)
  scale_fill_manual(values = c("2016" = "lightblue", "2019" = "seagreen")) # custom colors (fill)

```

## i. Write a caption for your figure

**Figure 1. Predicted Swift Parrot next box occupancy by distance from forest edge (m) and season.**

The figure shows predicted probabilities of Swift Parrot occupancy in nest boxes as a function of distance from forest edge (m) for two breeding seasons (2016 and 2019). The predictions were generated from the best fitting model according to Akaike's Information Criterion (AIC) and are visualized through a binomial generalized linear model, using ggpredict and ggplot. Shaded ribbons represent 95% confidence intervals and raw data is overlayed as jittered points. Seasons are differentiated by color with 2019 in green and 2016 in light blue.
**Reference Data:** Stojanovic, Dejan et al. (2021). *Do nest boxes breed the target species or its competitors? A case study of a critically endangered bird* [Dataset]. Dryad. https://doi.org/10.5061/dryad.83bk3j9sb

## j. Calculate model predictions 

```{r}
# calculate model predictions
model_predict <- ggpredict(model1, terms = c("edge_distance [0,900]", "season"))
model_predict
```

## k. Interpret your results

# Problem 4. Affective and exploratory visualizations

## a. Comparing visualizations

## b. Sharing your affective visualization


